A Tale of Two Transects

### 2020-02-11 writing/transect_paper.Rmd
move more code out of the paper into the local drake plan

base.dir <- regmatches(getwd(), regexpr(".*oueme-fungi-transect", getwd()))
if (opts_chunk$get("dev") == "tikz") {
  # styles$Expression <- styles$LaTeX.Expression
  # styles$Name <- styles$LaTeX.Name
  # styles$Short.Name <- styles$LaTeX.Short.Name
  # styles$Unit <- paste0("\\si{", styles$LaTeX.Unit, "}")
}
write_bib(c("base", "phyloseq", "gstat"), file = "R.bib", tweak = TRUE)
cache = drake_cache(here(".light"))
light_plan <- readRDS(here("data/plan/drake_light.rds"))
theme_set(theme_bw())

@@

k_or_M <- function(x, ..., .sep = " ", .fun = format) {
  case_when(
    abs(x) > 1e6 ~ paste(.fun(x / 1e6, ...), "M", sep = .sep),
    abs(x) > 1e3 ~ paste(.fun(x / 1e3, ...), "k", sep = .sep),
    is.na(x) ~ NA_character_,
    TRUE ~ paste(.fun(x, ...), " ", sep = .sep)
  )
}

percent <- function(x, ...) {
  paste0(formatC(as.numeric(x) * 100, ...), "%")
}

text_list <- function(x, ...) {
  x = format(x, ...)
  glue::glue_collapse(x, sep = ", ", last = ", and ")
}
```

@@

We attempted to denoise both long and short amplicons using `DADA2` version `r packageVersion("dada2")` [@callahan2016; @callahan2019] according to the steps outlined in  the supplementary information in @callahan2019.
However, only `r region_table$ASVs[region_table$region=="long"]` amplicon sequence variants (ASVs) were obtained for the long amplicons, representing `r round(100*region_table$reads[region_table$region == "long"]/as.integer(gsub(" ", "", demuxcounts["pb_500"])), 1)`% of the trimmed reads.
We believe that this poor performance was due to a combination of long read length and low sequencing depth relative to community diversity.
The `DADA2` algorithm requires that the seed sequence of each ASV be represented by at least two error-free reads [@callahan2016].
If sequencing errors are uniformly distributed, then the probability that a given read will be error-free is $(1-\epsilon)^L$, where $\epsilon$ is the sequencing error rate and $L$ is the read length in base pairs.
Then the number of reads of a given sequence that would be required to obtain two error-free reads in expectation is $2/(1-\epsilon)^L$.
For the combination of long reads (median $L=`r demuxlength['pb_500']` \text{ bp}$ after trimming) and moderate error rate (mean $\epsilon= `r format(demuxqual['pb_500'], digits = 4)`$ based on ccs quality scores) for the long amplicon in this study, the expected number of reads required to achieve two error-free reads is `r format(as.integer(round(2/(1 - demuxqual['pb_500'])^demuxlength['pb_500'])), big.mark = " ")`.
Given the high diversity relative to sequencing depth in this study (`r region_table$ASVs[region_table$seq_run == "pb_483" & region_table$region == "short"]` ASVs based on PacBio short amplicons, `r demuxcounts["pb_500"]` trimmed long amplicon reads), this requirement could not have been met for the long amplicons except by the most abundant sequences.
In comparison, the equivalent requirement for the short amplicon ($L=`r demuxlength['pb_483']` \text{ bp}$, $\epsilon=`r demuxqual['pb_483']`$) is only `r round(2/(1 - demuxqual['pb_483'])^demuxlength['pb_483'], 1)` reads.
We therefore developed a new workflow to assemble ASVs from the long amplicons, as follows:

@@

The 5.8S region was located in `r text_list(bioin)

@@

```{r bioinfo}
#TODO add filtered reads (currently inflated numbers for non-Illumina)
#TODO add Illumina raw reads
#TODO asterisks for mappde ASVs
loadd(bioinf_table, cache = cache)
options(knitr.kable.NA = '–')
bioinf_header <- tibble(name = names(bioinf_table)) %>%
  separate(name, into = c("tech", "amplicon", "type"), sep = "_") %>%
  mutate(tech = paste(tech, plyr::mapvalues(tech, datasets$tech, datasets$machine, FALSE)))
bioinf_table %>%
  set_names(gsub(".*_", "", names(.))) %>%
  kable(
    booktabs = TRUE,
    caption = "(ref:readcounts)",
    align = "rrrrrr",
    format.args = list(big.mark = " "),
    linesep = c(rep("", 3), rep("\\addlinespace", 2), rep("", 4)),
    escape = FALSE
    ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      bioinf_header$amplicon[c(TRUE, FALSE)] %>% set_names(rep(2, length(.)), .)
    )
  ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rle(bioinf_header$tech) %$% set_names(lengths, values)
    )
  )
```

@@

```{r venn-table-asv}
readd(venn_ASV, cache = cache) %>%
venn_table(ASVs, "(ref:venn-table-asv)")
```

@@

```{r venn-table-otu}
readd(venn_OTU, cache = cache) %>%
venn_table(OTUs, "(ref:venn-table-otu)")
```

@@

```{r venn, fig.cap = "(ref:venn)", out.height = "30%", fig.show="hold"}
readd(vennplot_ASV, cache = cache) %>% {
  plot(
    select(., set, ASVs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
    legend = list(
      labels = c("PacBio Long", "PacBio Short", "Ion Torrent", "Illumina")
    ),
    quantities = .$label,
    main = "ASV"
  )
}

readd(vennplot_OTU, cache = cache) %>% {
  plot(
    select(., set, OTUs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
    legend = FALSE,
    labels = FALSE,
    quantities = .$label,
    main = "OTU"
  )
}

@@

(ref:readcap) Comparison between read numbers for different sequencing strategies, by ASV (left) and 97% OTU (right). ASVs/OTUs which were detected by one sequencing strategy but not the other are plotted as tick marks along the axes. Dashed line represents a constant ratio of read numbers. The blue line is a LOESS smooth of the data, with associated uncertainty in grey shading. $R^2$ value displayed is for log-transformed non-zero read numbers.

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
loadd(
  big_table,
  multi_table,
  comparisons,
  cache = cache
)

compR2 <- function(x, y, type) {
  xamp <- plyr::mapvalues(
    x,
    datasets$seq_run,
    datasets$amplicon,
    FALSE
  )
  yamp <- plyr::mapvalues(
    y,
    datasets$seq_run,
    datasets$amplicon,
    FALSE
  )
  x <- plyr::mapvalues(
    x,
    datasets$seq_run,
    paste(
      datasets$tech,
      datasets$machine),
    FALSE
  )
  y <- plyr::mapvalues(
    y,
    datasets$seq_run,
    paste(
      datasets$tech,
      datasets$machine),
    FALSE
  )
  formatC(
    filter(
      comparisons,
      x == x_var,
      xamp == x_amplicon,
      y == y_var,
      yamp == y_amplicon,
      type == !!type
    )$r.squared,
    digits = 2,
    format = "f")
}

read_comparison <- function(multi_table, comparisons, type) {
  multi_table <- filter(multi_table, type == !!type)
  comparisons <- filter(comparisons, type == !!type)
  p <- multi_table %>%
    filter(x > 0, y > 0) %>%
    ggplot(aes(y = y, x = x)) +
    geom_point(alpha = 0.2, shape = 1) +
    scale_x_log10(
      breaks = c(1, 1000, 1000000),
      labels = c("1", "1k", "1M"),
      minor_breaks = c(10, 100, 10000, 100000),
      limits = c(1, 1e6)
    ) +
    scale_y_log10(
      breaks = c(1, 1000, 1000000),
      labels = c("1", "1k", "1M"),
      minor_breaks = c(10, 100, 10000, 100000),
      limits = c(1, 1e6)
    ) +
    coord_equal() +
    xlab(NULL) +
    ylab(NULL) +
    geom_rug(aes(y = y), sides = "l", alpha = 0.2,
             data = multi_table %>% filter(x == 0, y > 0)) +
    geom_rug(aes(x = x), sides = "b", alpha = 0.2,
             data = multi_table %>% filter(y == 0, x > 0)) +
    ggnomics::facet_nested(y_amplicon + y_var ~ x_amplicon + x_var, switch = "both", space = "free", nest_line = TRUE) +
    stat_smooth(method = "loess", formula = y ~ x) +
    geom_abline(aes(intercept = `(Intercept)`, slope = 1),
                linetype = 2L, alpha = 0.5, data = comparisons) +
    geom_text(
      aes(label = label),
      data = comparisons,
      x = 6,
      y = 0.7,
      color = "black",
      hjust = 1,
      parse = TRUE,
      inherit.aes = FALSE
    ) +
    theme(strip.placement = "outside", strip.background = element_blank())
  g <- ggplotGrob(p)
  g$grobs[g$layout$name %in% c("panel-1-2", "panel-1-3", "panel-2-3")] <- NULL
  g$layout <- g$layout[!(g$layout$name %in% c("panel-1-2", "panel-1-3", "panel-2-3")),]
  g
}
ggpubr::ggarrange(
  read_comparison(multi_table, comparisons, type = "ASV"),
  read_comparison(multi_table, comparisons, type = "OTU"),
  ncol = 1,
  labels = "auto",
  label.x = 0.9
)
```

The number of ASVs and OTUs shared between the different sequencing strategies are shown in Tables \@ref(tab:venn-table-asv) and \@ref(tab:venn-table-otu), respectively, and graphically in figure \@ref(fig:venn).
ASVs shared between all datasets represented `r text_list(percent(venn_ASV["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` of the reads for the long and short PacBio datasets, Illumina dataset, and Ion Torrent dataset, respectively.
These fractions increased to `r text_list(percent(venn_OTU["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` when differences at the intra-species scale were removed by clustering the ASVs into 97% OTUs.
In particular, `r text_list(percent(colSums(mutate_all(venn_OTU[c("0111", "1111"), c("frac_pb_483", "frac_SH-2257", "frac_is_057")], as.numeric))))` of reads in the PacBio, Illumina, and Ion Torrent short-amplicon datasets belonged to OTUs shared between all three datasets.
In contrast, `r percent(venn_ASV["1000", "frac_pb_500"])` of reads in the long PacBio dataset belonged to ASVs which were unique to that dataset, and the fraction only reduced to `r percent(venn_OTU["1000", "frac_pb_500"])` after OTU clustering.

Figure \@ref(fig:pb-vs-it) shows the correspondence between the read count for different ASVs/OTUs in the different technologies, where shared ASVs/OTUs are plotted as circles, and unshared OTUs are plotted on the margins.
In all cases, the read counts for shared ASVs/OTUs were correlated ($R^2 = compR2("SH-2257", "is_057", "OTU")`$).
Correlations between read counts for the three technologies using the short amplicon library were increased by OTU clustering
(`r compR2("pb_483", "SH-2257", "ASV")` to `r compR2("pb_483", "SH-2257", "OTU")`,
`r compR2("pb_483", "is_057", "ASV")` to `r compR2("pb_483", "is_057", "OTU")`, and
`r compR2("SH-2257", "is_057", "ASV")` to `r compR2("SH-2257", "is_057", "OTU")`,
for PacBio vs. Illumina, PacBio vs. Ion Torrent, and Illumina vs. Ion Torrent, respectively),
but this was not, or only weakly, the case for correlations between the long amplicon library and short amplicon library
(`r compR2("pb_500", "pb_483", "ASV")` to `r compR2("pb_500", "pb_483", "OTU")`,
`r compR2("pb_500", "SH-2257", "ASV")` to `r compR2("pb_500", "SH-2257", "OTU")`, and
`r compR2("pb_500", "is_057", "ASV")` to `r compR2("pb_500", "is_057", "OTU")`, for PacBio long amplicon reads vs PacBio, Illumina, and Ion Torrent short reads, respectively).

@@

```{r tax-chart, fig.cap = "(ref:tax-chart)"}
readd(taxon_reads, cache = cache) %>%
  pivot_longer(names_to = "rank", cols = c("kingdom", "phylum", "class", "order", "family", "genus"), values_to = "taxon") %>%
  group_by(amplicon, tech, Algorithm, reference, rank, ID = !is.na(taxon)) %>%
  summarize(reads = sum(reads), ASVs = n()) %>%
  group_by(amplicon, tech, Algorithm, reference, rank) %>%
  mutate(ASVs = ASVs / sum(ASVs)) %>%
  ungroup() %>%
  filter(ID) %>%
  mutate_at("rank", factor, levels = c("kingdom", "phylum", "class", "order", "family", "genus")) %>%
  rename(Reads = reads) %>%
  mutate_at("reference", as.character) %>%
  mutate_at("reference", replace_na, "All") %>%
  mutate(
    reference = map2(reference, amplicon, ~ if (.x == "All") c("Unite", "Warcup", if (.y == "Long") "RDP" else NULL) else .)
  ) %>%
  unnest(reference) %>%
  mutate_at("reference", factor, levels = c("Unite", "Warcup", "RDP")) %>%
  pivot_longer(cols = c("Reads", "ASVs"), names_to = "type", values_to = "frac") %>%
  ggplot(aes(x = rank, y = frac, ymax = frac, group = Algorithm, fill = Algorithm)) +
  ggnomics::facet_nested(
    tech + amplicon ~ type + reference,
    resect = unit(1, "mm"),
    nest_line = TRUE,
    bleed = FALSE
  ) +
  # geom_ribbon(ymin = 0, alpha = 0.5) +
  # geom_line(aes(color = Algorithm)) +
  geom_col(position = "dodge") +
  # scale_fill_brewer(type = "qual", palette = 2) +
  ylab("Fraction of reads assigned") +
  xlab(NULL) +
  theme(axis.text.x = element_text(vjust = 0.5, angle = 90),
        legend.direction = "horizontal",
        legend.position = "bottom",
        strip.background = element_blank())
```

(ref:taxa) Heat tree summarizing taxonomic composition of soil community.
Color and size of nodes represent the fractional read abundance of all ASVs belonging to that taxonomic group.
Color and thickness of branches represent the fractional ASV richness of that taxonomic group.
Both measures are a mean across sequencing runs, reference databases, and taxonomic assignment algorithms.
Diverse but relatively rare groups are thus shown as small, light-colored nodes on relatively thick, darker-green-colored branches (e.g., Dothideomycetes), while common but relatively non-diverse groups are shown as large nodes on relatively thin branches (e.g., Russulales).
Groups which are not represented by at least 1% of reads or 1% of ASVs in any dataset are collapsed into nodes labeled "...".
ASVs which could not be further assigned are labelled "?".

```{r taxa, fig.cap = "(ref:taxa)", out.width="100%", fig.width = 8}
fungi_read <- readd(taxdata, cache = cache) %>%
  taxa::filter_taxa(
    taxon_names == "Fungi",
    reassign_obs = FALSE,
    reassign_taxa = FALSE
    ) %>%
  extract2("data") %>%
  extract2("tax_read") %>%
  select(ends_with("PHYLOTAX")) %>%
  set_names(names(.) %>% str_extract("(PacBio|Illumina|Ion.Torrent)_(Long|Short)")) %>%
  unlist()

fungi_asv <- readd(taxdata, cache = cache) %>%
  taxa::filter_taxa(
    taxon_names == "Fungi",
    reassign_obs = FALSE,
    reassign_taxa = FALSE
    ) %>%
  extract2("data") %>%
  extract2("tax_asv") %>%
  select(ends_with("PHYLOTAX")) %>%
  set_names(names(.) %>% str_extract("(PacBio|Illumina|Ion.Torrent)_(Long|Short)")) %>%
  unlist()

ecm_counts <- readd(taxon_reads, cache = cache) %>%
  filter(Algorithm == "PHYLOTAX") %>%
  group_by(seq_run) %>%
  mutate(ASVs = 1/n()) %>%
  filter(ECM %in% c("Possible ECM", "Probable ECM", "Highly Probable ECM")) %>%
  summarize_at(c("ASVs", "reads"), sum)

@@

set.seed(2)
theme <- theme_update(panel.border = element_blank())
readd(taxdata, cache = cache) %>%
  metacoder::heat_tree(
    layout = "davidson-harel",
    # initial_layout = "reingold-tilford",
    node_size = rowMeans(.$data$tax_read[,-1]),
    node_color = rowMeans(.$data$tax_read[,-1]),
    node_label = replace_na(taxon_names, "unknown"),
    node_size_range = c(0.002, .035),
    node_size_axis_label = "Read abundance",
    node_color_axis_label = "Read abundance",
    node_label_size_range = c(0.01, 0.03),
    edge_size = rowMeans(.$data$tax_asv[,-1]),
    edge_size_trans = "linear",
    edge_size_axis_label = "ASV count",
    edge_size_range = c(0.001, 0.03),
    edge_color = rowMeans(.$data$tax_asv[,-1]),
    edge_color_axis_label = "ASV count",
    # aspect_ratio = 3/2,
    make_node_legend = FALSE,
    make_edge_legend = FALSE,
    output_file = here("temp/heattree.pdf")
  )
theme_set(theme)
```

@@

```{r taxon-plot}
taxon_plot <- function(.data, rank, ..., y = reads, cutoff = NULL) {
  rank <- enquo(rank)
  y <- enquo(y)
  ranks <- c("kingdom", "phylum", "class", "order", "family", "genus")
  .data <- .data %>%
    group_by(reference, Algorithm, tech, amplicon) %>%
    mutate(ASVs = n()) %>%
    ungroup() %>%
    filter(...) %>%
    arrange_at(ranks) %>%
    mutate_at(
      ranks,
      ~ factor(
        .,
        levels = c(NA, "other", discard(unique(.), is.na)),
        exclude = NULL
      )
    ) %>%
    group_by(reference, Algorithm, tech, amplicon, !!rank) %>%
    summarize(reads = sum(reads), ASVs = n()/max(ASVs)) %>%
    ungroup()

  if (!is.null(cutoff)) {
    .data <- group_by(.data, !!rank) %>%
      group_map(

        ~ if (all(pull(.x, !!y) < cutoff)) {
          mutate(.x, !!rank := factor("other", levels = levels(!!rank)))
        } else {
          .x
        },
        keep = TRUE
      ) %>%
      bind_rows() %>%
      group_by(reference, Algorithm, !!rank, tech, amplicon) %>%
      summarize(reads = sum(reads), ASVs = sum(ASVs)) %>%
      ungroup()
  }

  rank_label <- as_label(rank)

  .data <- mutate(.data, !!rank := fct_drop(!!rank))
  vals <- levels(pull(.data, !!rank))
  # if ("other" %in% vals) vals <- c("other", vals) %>% magrittr::extract(!duplicated(.))
  # if (any(is.na(vals))) vals <- c(NA, vals) %>% magrittr::extract(!duplicated(.))

  if (rank_label == str_to_lower(rank_label)) rank_label <- str_to_title(rank_label)
  y_label <- as_label(y)
  if (y_label == str_to_lower(y_label)) y_label <- str_to_title(y_label)
  .data <- .data %>%
    mutate(tech = fct_relabel(tech, ~ paste(., plyr::mapvalues(., datasets$tech, datasets$machine, FALSE))
    ))
  ggplot(.data, aes(x = Algorithm, y = !!y, fill = !!rank)) +
    geom_bar(position = "stack", stat = "identity", color = "white", size = 0.2) +
    ggnomics::facet_nested(
      cols = vars(tech, amplicon, reference),
      scales = "free_x",
      space = "free_x",
      nest_line = TRUE,
      bleed = FALSE,
      resect = unit(1, "mm")
      ) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
          strip.background = element_blank(),
          panel.spacing = unit(3, "pt")) +
    scale_fill_discrete(
      breaks = vals,
      labels = replace_na(as.character(vals), "unidentified"),
      name = rank_label
    ) +
    ylab(paste("Fraction of", y_label))

}

@@

```{r heattree-db-compare, fig.cap = "(ref:heattree-db-compare)", include = FALSE}
taxdata2 <- readd(taxdata, cache = cache)
# taxdata2$data$asv_table <- metacoder::calc_obs_props(taxdata2, "tax_data", cols = asv_cols)
# taxdata2$data$read_table <- metacoder::calc_obs_props(taxdata2, "tax_data", cols = read_cols)
# taxdata2$data$tax_asv <- metacoder::calc_taxon_abund(taxdata2, "asv_table", cols = asv_cols)
# taxdata2$data$tax_read <- metacoder::calc_taxon_abund(taxdata2, "read_table", cols = read_cols)
taxdata2$data$diff_read <- metacoder::compare_groups(
  taxdata2,
  data = "tax_read",
  cols = names(taxdata2$data$tax_read) %>%
    keep(str_detect, "reads_PacBio_Long_(Unite|RDP)"),
  groups = names(taxdata2$data$tax_read) %>%
    keep(str_detect, "reads_PacBio_Long_(Unite|RDP)") %>%
    str_replace("reads_.+_.+_(.+)_.+", "\\1"),
  func = function(abund1, abund2) {
    list(read_ratio = log10(mean(abund1) / mean(abund2)))
  }
  )
taxdata2$data$diff_asv <- metacoder::compare_groups(
  taxdata2,
  data = "tax_asv",
  cols = names(taxdata2$data$tax_asv) %>% keep(startsWith, "ASVs_"),
  groups = names(taxdata2$data$tax_asv) %>%
    keep(startsWith, "ASVs_") %>%
    str_replace("ASVs_.+_(.+)_.+_.+", "\\1"),
  func = function(abund1, abund2) {
    list(asv_ratio = log10(mean(abund1) / mean(abund2)))
  }
)

@@

full_ecdf <- readd(reads_table, cache = cache) %>%
  dplyr::filter(region  %in% c("long", "short")) %>%
  left_join(select(datasets, seq_run, amplicon, tech, machine), by = "seq_run") %>%
  mutate_at("tech", factor, levels = c("PacBio", "Illumina", "Ion Torrent")) %>%
  dplyr::group_by(seq_run) %>%
  dplyr::mutate(reads = reads / sum(reads))

>>>>

base.dir <- regmatches(getwd(), regexpr(".*oueme-fungi-transect", getwd()))
if (opts_chunk$get("dev") == "tikz") {
  # styles$Expression <- styles$LaTeX.Expression
  # styles$Name <- styles$LaTeX.Name
  # styles$Short.Name <- styles$LaTeX.Short.Name
  # styles$Unit <- paste0("\\si{", styles$LaTeX.Unit, "}")
}
write_bib(c("base", "phyloseq", "gstat"), file = "R.bib", tweak = TRUE)

# use the "light" plan cache instead of the default cache at .drake
cache <- drake_cache(here(".light"))
readd <- function(..., cache = drake_cache(here(".light"))) {
  force(cache)
  drake::readd(..., cache = cache)
}
loadd <- function(..., cache = drake_cache(here(".light"))) {
  force(cache)
  drake::loadd(..., cache = cache, envir = parent.frame())
}

light_plan <- readRDS(here("data/plan/drake_light.rds"))
theme_set(theme_bw())

@@

source(here(file.path("scripts", "output_functions.R")))
```

@@

We attempted to denoise both long and short PacBio amplicons using `DADA2` version `r packageVersion("dada2")` [@callahan2016; @callahan2019] according to the steps outlined in  the supplementary information in @callahan2019.
However, only `r readd(region_table)$ASVs[readd(region_table)$region=="long"]` amplicon sequence variants (ASVs) were obtained for the long amplicons, representing `r percent(readd(region_table)$reads[readd(region_table)$region == "long"]/as.integer(gsub(" ", "", readd(demuxcounts)["pb_500"])), format = "d")` of the trimmed reads.
We believe that this poor performance was due to a combination of long read length and low sequencing depth relative to community diversity.
The `DADA2` algorithm requires that the seed sequence of each ASV be represented by at least two error-free reads [@callahan2016].
If sequencing errors are uniformly distributed, then the probability that a given read will be error-free is $(1-\epsilon)^L$, where $\epsilon$ is the sequencing error rate and $L$ is the read length in base pairs.
Then the number of reads of a given sequence that would be required to obtain two error-free reads in expectation is $2/(1-\epsilon)^L$.
For the combination of long reads (median $L=`r demuxlength['pb_500']` \text{ bp}$ after trimming) and moderate error rate (mean $\epsilon= `r format(demuxqual['pb_500'], digits = 4)`$ based on ccs quality scores) for the long amplicon in this study, the expected number of reads required to achieve two error-free reads is `r format(as.integer(round(2/(1 - demuxqual['pb_500'])^demuxlength['pb_500'])), big.mark = " ")`.
Given the high diversity relative to sequencing depth in this study (`r region_table$ASVs[region_table$seq_run == "pb_483" & region_table$region == "short"]` ASVs based on PacBio short amplicons, `r demuxcounts["pb_500"]` trimmed long amplicon reads), this requirement could not have been met for the long amplicons except by the most abundant sequences.
In comparison, the equivalent requirement for the short amplicon ($L=`r demuxlength['pb_483']` \text{ bp}$, $\epsilon=`r demuxqual['pb_483']`$) is only `r round(2/(1 - demuxqual['pb_483'])^demuxlength['pb_483'], 1)` reads.
We therefore developed a new workflow to assemble ASVs from the long amplicons, as follows:

@@

@@

```{r bioinfo}
#TODO add filtered reads (currently inflated numbers for non-Illumina)
#TODO asterisks for mapped ASVs
# loadd(bioinf_table)
options(knitr.kable.NA = '–')

readd(bioinf_table) %>%
  set_names(gsub(".*_", "", names(.))) %>%
  kable(
    booktabs = TRUE,
    caption = "(ref:readcounts)",
    align = "rrrrrr",
    format.args = list(big.mark = " "),
    linesep = c(rep("", 3), rep("\\addlinespace", 2), rep("", 4)),
    escape = FALSE
    ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      readd(bioinf_header)$amplicon[c(TRUE, FALSE)] %>% set_names(rep(2, length(.)), .)
    )
  ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rle(readd(bioinf_header)$tech) %$% set_names(lengths, values)
    )
  )
```

@@

```{r venn-table-asv}
readd(venn_ASV) %>%
venn_table(ASVs, "(ref:venn-table-asv)")
```

@@

```{r venn-table-otu}
readd(venn_OTU) %>%
venn_table(OTUs, "(ref:venn-table-otu)")
```

@@

```{r venn, fig.cap = "(ref:venn)", out.height = "30%", fig.show="hold"}
readd(vennplot_ASV) %>% {
  plot(
    select(., set, ASVs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
    legend = list(
      labels = c("PacBio Long", "PacBio Short", "Ion Torrent", "Illumina")
    ),
    quantities = .$label,
    main = "ASV"
  )
}

readd(vennplot_OTU) %>% {
  plot(
    select(., set, OTUs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
    legend = FALSE,
    labels = FALSE,
    quantities = .$label,
    main = "OTU"
  )
}

@@

(ref:readcap) Comparison between read numbers for different sequencing strategies, by ASV (top) and 97% OTU (bottom). ASVs/OTUs which were detected by one sequencing strategy but not the other are plotted as tick marks along the axes. Dashed line represents a constant ratio of read numbers. The blue line is a LOESS smooth of the data, with associated uncertainty in grey shading. $R^2$ value displayed is for log-transformed non-zero read numbers.

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
ggpubr::ggarrange(
  read_comparison(
    readd(multi_table),
    readd(comparisons),
    type = "ASV"
  ),
  read_comparison(
    readd(multi_table),
    readd(comparisons),
    type = "OTU"
  ),
  ncol = 1,
  labels = "auto",
  label.x = 0.9
)
```

The number of ASVs and OTUs shared between the different sequencing strategies are shown in Tables \@ref(tab:venn-table-asv) and \@ref(tab:venn-table-otu), respectively, and graphically in figure \@ref(fig:venn).
ASVs shared between all datasets represented `r text_list(percent(readd(venn_ASV)["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` of the reads for the long and short PacBio datasets, Illumina dataset, and Ion Torrent dataset, respectively.
These fractions increased to `r text_list(percent(readd(venn_OTU)["1111", c("frac_pb_500", "frac_pb_483", "frac_SH-2257", "frac_is_057")]))` when differences at the intra-species scale were removed by clustering the ASVs into 97% OTUs.
In particular, `r text_list(percent(colSums(mutate_all(readd(venn_OTU)[c("0111", "1111"), c("frac_pb_483", "frac_SH-2257", "frac_is_057")], as.numeric), na.rm = TRUE)))` of reads in the PacBio, Illumina, and Ion Torrent short-amplicon datasets belonged to OTUs shared between all three datasets.
In contrast, `r percent(readd(venn_ASV)["1000", "frac_pb_500"])` of reads in the long PacBio dataset belonged to ASVs which were unique to that dataset, and the fraction only reduced to `r percent(readd(venn_OTU)["1000", "frac_pb_500"])` after OTU clustering.

Figure \@ref(fig:pb-vs-it) shows the correspondence between the read count for different ASVs/OTUs in the different technologies, where shared ASVs/OTUs are plotted as circles, and unshared OTUs are plotted on the margins.
In all cases, the read counts for shared ASVs/OTUs were correlated ($R^2 = `r compR2("SH-2257", "is_057", "OTU")`$).
Correlations between read counts for the three technologies using the short amplicon library were increased by OTU clustering
(`r compR2("pb_483", "SH-2257", "ASV")` to `r compR2("pb_483", "SH-2257", "OTU")`,
`r compR2("pb_483", "is_057", "ASV")` to `r compR2("pb_483", "is_057", "OTU")`, and
`r compR2("SH-2257", "is_057", "ASV")` to `r compR2("SH-2257", "is_057", "OTU")`,
for PacBio vs. Illumina, PacBio vs. Ion Torrent, and Illumina vs. Ion Torrent, respectively),
but this was not, or only weakly, the case for correlations between the long amplicon library and short amplicon library
(`r compR2("pb_500", "pb_483", "ASV")` to `r compR2("pb_500", "pb_483", "OTU")`,
`r compR2("pb_500", "SH-2257", "ASV")` to `r compR2("pb_500", "SH-2257", "OTU")`, and
`r compR2("pb_500", "is_057", "ASV")` to `r compR2("pb_500", "is_057", "OTU")`, for PacBio long amplicon reads vs PacBio, Illumina, and Ion Torrent short reads, respectively).

@@

```{r tax-chart, fig.cap = "(ref:tax-chart)"}
readd(tax_chart)
```

(ref:taxa) Heat tree summarizing taxonomic composition of soil community.
Color and size of nodes represent the fractional read abundance of all ASVs belonging to that taxonomic group.
Color and thickness of branches represent the fractional ASV richness of that taxonomic group.
Both measures are a mean across sequencing runs, reference databases, and taxonomic assignment algorithms.
Diverse but relatively rare groups are thus shown as small, light-colored nodes on relatively thick, darker-green-colored branches (e.g., Dothideomycetes), while common but relatively non-diverse groups are shown as large nodes on relatively thin branches (e.g., Russulales).
Groups which are not represented by at least 1% of reads or 1% of ASVs in any dataset are collapsed into nodes labeled "*".
ASVs which could not be further assigned are labelled "?".

```{r taxa, fig.cap = "(ref:taxa)", out.width="100%", fig.width = 8}
fungi_read <- readd(taxdata) %>%
  taxa::filter_taxa(
    taxon_names == "Fungi",
    reassign_obs = FALSE,
    reassign_taxa = FALSE
    ) %>%
  extract2("data") %>%
  extract2("tax_read") %>%
  select(ends_with("PHYLOTAX")) %>%
  set_names(names(.) %>% str_extract("(PacBio|Illumina|Ion.Torrent)_(Long|Short)")) %>%
  unlist()

fungi_asv <- readd(taxdata) %>%
  taxa::filter_taxa(
    taxon_names == "Fungi",
    reassign_obs = FALSE,
    reassign_taxa = FALSE
    ) %>%
  extract2("data") %>%
  extract2("tax_asv") %>%
  select(ends_with("PHYLOTAX")) %>%
  set_names(names(.) %>% str_extract("(PacBio|Illumina|Ion.Torrent)_(Long|Short)")) %>%
  unlist()

ecm_counts <- readd(taxon_reads) %>%
  filter(Algorithm == "PHYLOTAX") %>%
  group_by(seq_run) %>%
  mutate(ASVs = 1/n()) %>%
  filter(ECM %in% c("Possible ECM", "Probable ECM", "Highly Probable ECM")) %>%
  summarize_at(c("ASVs", "reads"), sum)

@@

set.seed(2)
theme <- theme_update(panel.border = element_blank())
readd(taxdata) %>%
  metacoder::heat_tree(
    layout = "davidson-harel",
    # initial_layout = "reingold-tilford",
    node_size = rowMeans(.$data$tax_read[,-1]),
    node_color = rowMeans(.$data$tax_read[,-1]),
    node_label = replace_na(taxon_names, "unknown"),
    node_size_range = c(0.002, .035),
    node_size_axis_label = "Read abundance",
    node_color_axis_label = "Read abundance",
    node_label_size_range = c(0.01, 0.03),
    edge_size = rowMeans(.$data$tax_asv[,-1]),
    edge_size_trans = "linear",
    edge_size_axis_label = "ASV count",
    edge_size_range = c(0.001, 0.03),
    edge_color = rowMeans(.$data$tax_asv[,-1]),
    edge_color_axis_label = "ASV count",
    # aspect_ratio = 3/2,
    make_node_legend = FALSE,
    make_edge_legend = FALSE,
    output_file = here("temp/heattree.pdf")
  )
theme_set(theme)
```

@@

@@

```{r heattree-db-compare, fig.cap = "(ref:heattree-db-compare)", include = FALSE}
taxdata2 <- readd(taxdata)
# taxdata2$data$asv_table <- metacoder::calc_obs_props(taxdata2, "tax_data", cols = asv_cols)
# taxdata2$data$read_table <- metacoder::calc_obs_props(taxdata2, "tax_data", cols = read_cols)
# taxdata2$data$tax_asv <- metacoder::calc_taxon_abund(taxdata2, "asv_table", cols = asv_cols)
# taxdata2$data$tax_read <- metacoder::calc_taxon_abund(taxdata2, "read_table", cols = read_cols)
taxdata2$data$diff_read <- metacoder::compare_groups(
  taxdata2,
  data = "tax_read",
  cols = names(taxdata2$data$tax_read) %>%
    keep(str_detect, "reads_PacBio_Long_(Unite|RDP)"),
  groups = names(taxdata2$data$tax_read) %>%
    keep(str_detect, "reads_PacBio_Long_(Unite|RDP)") %>%
    str_replace("reads_.+_.+_(.+)_.+", "\\1"),
  func = function(abund1, abund2) {
    list(read_ratio = log10(mean(abund1) / mean(abund2)))
  }
  )
taxdata2$data$diff_asv <- metacoder::compare_groups(
  taxdata2,
  data = "tax_asv",
  cols = names(taxdata2$data$tax_asv) %>% keep(startsWith, "ASVs_"),
  groups = names(taxdata2$data$tax_asv) %>%
    keep(startsWith, "ASVs_") %>%
    str_replace("ASVs_.+_(.+)_.+_.+", "\\1"),
  func = function(abund1, abund2) {
    list(asv_ratio = log10(mean(abund1) / mean(abund2)))
  }
)

@@

full_ecdf <- readd(reads_table) %>%
  dplyr::filter(region  %in% c("long", "short")) %>%
  left_join(select(datasets, seq_run, amplicon, tech, machine), by = "seq_run") %>%
  mutate_at("tech", factor, levels = c("PacBio", "Illumina", "Ion Torrent")) %>%
  dplyr::group_by(seq_run) %>%
  dplyr::mutate(reads = reads / sum(reads))
